Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
just linebreaks

\conferenceinfo{POPL '14}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2014}

@@

\begin{abstract}
Proving program termination is typically done by finding a well-founded \emph{ranking function}
for the program states.
Existing termination provers typically find ranking functions
using either linear algebra or templates.  As such they are often restricted to
finding linear ranking functions over mathematical integers.  This class
of ranking functions is not large enough to prove termination for all terminating
programs, and furthermore a termination argument for a program operating on mathematical integers
does not always lead to a termination argument for the same program operating on
fixed-width machine integers.

We present a reduction from program \emph{termination} to program \emph{synthesis}.
This reduction allows us to generate nonlinear, lexicographic ranking functions that
are correct for fixed-width machine arithmetic and floating-point arithmetic.
We use this reduction to build a sound and complete procedure for the termination
of fixed-width and floating-point arithmetic programs.
\end{abstract}

@@

As Ramsey-based approaches are limited by the state of the art in safety checking, 
recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments. 
This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any particular syntactic form, and do not use templates).

While the summary  in Figure~\ref{fig:handletable} supports the observation that the majority of existing termination analyses are designed for
linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
that bit-vector semantics and integer semantics give rise to the same termination behaviour. Thus, most of the techniques treat
bit-vectors and floats as mathematical integers and reals respectively 
\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector semantics to be identical to integer semantics,
these techniques ignore the wrap-around behaviour caused by overflows,
which can be unsound.
In Section~\ref{sec:motivation}, we show that integers and bit-vectors exhibit
incomparable behaviours with respect to termination, i.e.
programs that terminate for integers may \emph{not} terminate for bit-vectors and
vice-versa.
Thus, abstracting bit-vectors to integers may give rise to 
{\em unsound} and {\em incomplete} analyses.
%design decisions
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
lexicographic, non-linear 
ranking functions supported by non-linear inductive invariants for loops with 
arbitrary guards and transitions over bit-vectors and floats.
It also computes non-linear recurrence sets, which comprise non-termination proofs.
Our encoding uniformly multiple loops and nested loops, without having to enumerate
lassos.
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

@@

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
This means that our procedure is not directly dependent on the structure of the analysed loop, e.g. number of variable, number of lassos, 
but on the Kolmogorov complexity of its minimal termination argument. 
We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
programmers tend to write programs that have relatively short termination proofs. 
We empirically show that this claim holds in practice, causing our technique to perform well.

@@

 The main contributions of our work can be summarised as follows:

\begin{itemize}
\item We designed a bit-level accurate technique for computing ranking functions that correctly accounts for the wrap-around behavior
caused by under- and overflows in bit-vector and floating point arithmetic. Our technique is not restricted to finding linear ranking functions,
but can also compute lexicographic non-linear  ones.

@@

\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith} 
Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
over fixed-width binary words, otherwise known as bit-vectors.  For the remainder of this section, we will say that
the bit-vectors we are working with are $k$-bits wide, which means that each word can hold one of $2^k$ bit patterns.
Typical values for $k$ are 32 and 64.

@@

The problem of program verification can be reduced to the problem
of finding solutions to a second-order constraint \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
Our intention is to apply this approach to termination analysis. 
In this section we show how several variations of both the termination and the non-termination problem can be defined in second-order logic.  
%The second-order unknowns in this constraint are the unknown program invariants
%that are inductive and strong enough to prove the desired assertions. In this section we describe the conversion of programs to constraints

@@

We will begin our discussion by showing how to encode the (non-)termination of program consisting of a single loop 
in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop body.
For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

Only for certain values of \texttt{y} the loop is non-terminating.
In order to restrict the values that may be assigned to \texttt{y}, we need an assumption associated with the non-deterministic choice. 
However, such an assumption may cause a reachable state at the non-deterministic assignment to not have any successor. In such a case
the termination formula would be satisfiable, although the execution halts. This situation can be avoided by checking the existence of 
at least one successor for each reachable state at the non-deterministic assignment. Although we did put a reasonable amount of effort into it, 
we failed to come up with a formulation for the non-termination of such non-deterministic transition relations 
inside the synthesis fragment.
%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.

@@

\noindent{\bf Nested loops.}
For brevity reasons, we illustrate our approach to handling nested loops for only two loops, but the approach can be easily generalised to any number of nested loops:
\begin{lstlisting}[mathescape=true]
while ($G_1$) { 
  $P_1$ 
  while ($G_2$) { $T$ }
  $P_2$
}
\end{lstlisting}

@@

As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity. 
\fi
%
%

@@

The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$ which satisfy the partial specification:
\[
 \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
\]

@@

\noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}
To generate candidate (non-)termination proofs in the synthesise phase of the refinement loop,
we make use of genetic programming (GP) \cite{langdon:fogp} as an alternative to model checking and exhaustive search.  
Genetic programming provides an intelligent and adaptive way of searching through the space of computer programs
for an individual that is highly fit in solving the problem at hand
by using the Darwinian principle of survival and reproduction of the fittest.
Thus, the program that emerges from the genetic programming paradigm is a consequence of fitness.

@@

In the setting of program synthesis, genetic programming can be very usefully combined
with the concept of incremental evolution.  The idea is that in order to evolve a program
that is adequate for some difficult task, it is often helpful to first evolve a population
that as an ensemble is good at some simpler variation of the task.  This principle
leads us to the observation that each round of the CEGIS loop can be thought of
as creating a more difficult problem for the GP synthesiser to solve.  Therefore,
if GP finds a candidate program which is correct for the $k$ inputs so far observed,
but is not correct for the full specification, that candidate program came from a population
that was ``good'' for the specification restrict to those $k$ inputs.  The next time round
the CEGIS loop, the problem GP must solve is slightly harder -- any solution to this new
problem must be correct for the previous $k$ inputs and also whatever new input was added
by the previous iteration of {\sc verif}.  It therefore makes sense to begin the $k+1$st
round of evolution with the population at the end of the $k$th round.

@@

\section{Experiments}
We implemented our algorithm in the {\sc Headshot} tool, which uses the {\sc Kalashnikov} program synthesiser for termination.
We ran the tool on 40 benchmarks taken from the termination category of SVCOMP'15.
The experiments were performed on a 4-core 3.3GHz i5-2500k with 8GB of RAM.

@@

While the synthesis of termination arguments for linear programs over integers is indeed well-covered in the literature, 
there is very limited work for programs over machine integers.
Cook et al. present a method based on a reduction to Presburger
arithmetic, and a template-matching approach for predefined classes of
ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.
Similarly, the only work we are aware of that can compute non-linear ranking functions  
for imperative loops with polynomial guards and polynomial assignments
is \cite{DBLP:conf/vmcai/BradleyMS05}. However, this work extends only to polynomials.

@@

As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's theorem compute a set of local termination conditions that decrease as execution follows through the loop
and require expensive reachability analyses
\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
In an attempt to reduce the complexity of checking the validity of the termination argument, 
Cook et al present an iterative termination proving procedure that searches for 
lexicographic termination arguments \cite{DBLP:conf/tacas/CookSZ13}, 
whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.

>>>>

\conferenceinfo{POPL '15}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2015}

@@

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
not large enough to prove termination for all terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

We present a reduction from program \emph{termination} to program
\emph{synthesis}.  This reduction allows us to generate nonlinear,
lexicographic ranking functions that are correct for fixed-width machine
arithmetic and floating-point arithmetic.  We use this reduction to build a
sound and complete procedure for the termination of fixed-width and
floating-point arithmetic programs.
%
\end{abstract}

@@

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments.  This
means that our ranking functions may involve non-linearity and lexicographic
orders (we do not commit to any particular syntactic form, and do not use
templates).

While the summary  in Figure~\ref{fig:handletable} supports the observation
that the majority of existing termination analyses are designed for linear
programs and linear ranking functions, it also identifies another common
assumption made by state of the art termination provers: that bit-vector
semantics and integer semantics give rise to the same termination behaviour. 
Thus, most of the techniques treat bit-vectors and floats as mathematical
integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector semantics to be identical to integer semantics, these
techniques ignore the wrap-around behaviour caused by overflows, which can
be unsound.  In Section~\ref{sec:motivation}, we show that integers and
bit-vectors exhibit incomparable behaviours with respect to termination,
i.e.  programs that terminate for integers may \emph{not} terminate for
bit-vectors and vice-versa.  Thus, abstracting bit-vectors to integers may
give rise to {\em unsound} and {\em incomplete} analyses.
%
%design decisions
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
lexicographic, non-linear 
ranking functions supported by non-linear inductive invariants for loops with 
arbitrary guards and transitions over bit-vectors and floats.
It also computes non-linear recurrence sets, which comprise non-termination proofs.
Our encoding uniformly multiple loops and nested loops, without having to enumerate
lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

@@

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
%
This means that our procedure is not directly dependent on the structure of
the analysed loop, e.g.  number of variable, number of lassos, but on the
Kolmogorov complexity of its minimal termination argument.  We argue that
the length of a program's termination proof is indicative of how easy to
understand the program is, and thus programmers tend to write programs that
have relatively short termination proofs.  We empirically show that this
claim holds in practice, causing our technique to perform well.

@@

The main contributions of our work can be summarised as follows:
%
\begin{itemize}

\item We designed a bit-level accurate technique for computing ranking
functions that correctly accounts for the wrap-around behavior caused by
under- and overflows in bit-vector and floating point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

@@

\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith}

Physical computers have bounded storage, which means they are unable to
perform calculations on mathematical integers.  For example, if $A$ is the
Ackermann function and $G$ is Graham's number, a physical computer capable
of computing $A(G, G)$ would contain (much!) more matter than is believed to
exist in the universe.  Fortunately, it is rare for a programmer to need
such a large number and so modern computers do their arithmetic over
fixed-width binary words, otherwise known as bit-vectors.  For the remainder
of this section, we will say that the bit-vectors we are working with are
$k$-bits wide, which means that each word can hold one of $2^k$ bit
patterns.  Typical values for $k$ are 32 and 64.

@@

The problem of program verification can be reduced to the problem of finding
solutions to a second-order
constraint~\cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}. 
Our intention is to apply this approach to termination analysis.  In this
section we show how several variations of both the termination and the
non-termination problem can be defined in second-order logic.
%
%The second-order unknowns in this constraint are the unknown program invariants
%that are inductive and strong enough to prove the desired assertions. In this section we describe the conversion of programs to constraints

@@

We will begin our discussion by showing how to encode the (non-)termination
of program consisting of a single loop in the synthesis fragment.  For the
time being, a loop is a pair ${L= \langle G,T \rangle}$, where the states
$x$ satisfying the loop's guard are given by the predicate $G(x)$.  The body
of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop
body.  For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

Only for certain values of \texttt{y} the loop is non-terminating.  In order
to restrict the values that may be assigned to \texttt{y}, we need an
assumption associated with the non-deterministic choice.  However, such an
assumption may cause a reachable state at the non-deterministic assignment
to not have any successor.  In such a case the termination formula would be
satisfiable, although the execution halts.  This situation can be avoided by
checking the existence of at least one successor for each reachable state at
the non-deterministic assignment.  Although we did put a reasonable amount
of effort into it, we failed to come up with a formulation for the
non-termination of such non-deterministic transition relations inside the
synthesis fragment.
%
%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.

@@

\noindent{\bf Nested loops.}
%
For brevity reasons, we illustrate our approach to handling nested loops for
only two loops, but the approach can be easily generalised to any number of
nested loops:
%
\begin{lstlisting}[mathescape=true]
while ($G_1$) { 
  $P_1$ 
  while ($G_2$) { $T$ }
  $P_2$
}
\end{lstlisting}

@@

As a consequence of the aforementioned conjecture, we do not need to commit
to a specific form of the termination argument.  While other approaches,
e.g.  \cite{DBLP:conf/tacas/LeikeH14}, conduct independent searches for each
possible form of the ranking function with some of them futile whenever the
assumed constrained ranking function does not exist, we have a single search
that aims at finding the ranking function with the lowest Kolmogorov
complexity.
%
\fi
%
%

@@

The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
which satisfy the partial specification:
%
\[
 \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
\]

@@

\noindent{\bf Symbolic Bounded Model Checking} Another complete method for generating
candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
search, we must progressively increase the length of the solution we search for
in order to get a complete search procedure.
\\

\noindent {\bf Genetic Programming and Incremental Evolution} \label{sec:gp}
To generate candidate (non-)termination proofs in the synthesise phase of
the refinement loop, we make use of genetic programming
(GP)~\cite{langdon:fogp} as an alternative to model checking and exhaustive
search.  Genetic programming provides an intelligent and adaptive way of
searching through the space of computer programs for an individual that is
highly fit in solving the problem at hand by using the Darwinian principle
of survival and reproduction of the fittest.  Thus, the program that emerges
from the genetic programming paradigm is a consequence of fitness.

@@

In the setting of program synthesis, genetic programming can be very
usefully combined with the concept of incremental evolution.  The idea is
that in order to evolve a program that is adequate for some difficult task,
it is often helpful to first evolve a population that as an ensemble is good
at some simpler variation of the task.  This principle leads us to the
observation that each round of the CEGIS loop can be thought of as creating
a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
finds a candidate program which is correct for the $k$ inputs so far
observed, but is not correct for the full specification, that candidate
program came from a population that was ``good'' for the specification
restrict to those $k$ inputs.  The next time round the CEGIS loop, the
problem GP must solve is slightly harder -- any solution to this new problem
must be correct for the previous $k$ inputs and also whatever new input was
added by the previous iteration of {\sc verif}.  It therefore makes sense to
begin the $k+1$st round of evolution with the population at the end of the
$k$-th round.

@@

\section{Experiments}

We implemented our algorithm in the {\sc Headshot} tool, which uses the {\sc
Kalashnikov} program synthesiser for termination.  We ran the tool on 40
benchmarks taken from the termination category of SVCOMP'15.  The
experiments were performed on a 4-core 3.3\,GHz i5-2500k with 8\,GB of RAM.

@@

While the synthesis of termination arguments for linear programs over
integers is indeed well-covered in the literature, there is very limited
work for programs over machine integers.  Cook et al.  present a method
based on a reduction to Presburger arithmetic, and a template-matching
approach for predefined classes of ranking functions based on reduction to
SAT- and QBF-solving~\cite{DBLP:conf/tacas/CookKRW10}.  Similarly, the only
work we are aware of that can compute non-linear ranking functions for
imperative loops with polynomial guards and polynomial assignments
is~\cite{DBLP:conf/vmcai/BradleyMS05}.  However, this work extends only to
polynomials.

@@

As mentioned in Section~\ref{sec:intro}, approaches based on Ramsey's
theorem compute a set of local termination conditions that decrease as
execution follows through the loop and require expensive reachability
analyses~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
In an attempt to reduce the complexity of checking the validity of the
termination argument, Cook et al present an iterative termination proving
procedure that searches for lexicographic termination
arguments~\cite{DBLP:conf/tacas/CookSZ13}, whereas Kroening et al. 
strengthen the termination argument such that it becomes a transitive
relation~\cite{DBLP:conf/cav/KroeningSTW10}.
